Lesson 3
========================================================

***

### What to Do First?
Notes:

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
pf = read.csv('/Users/Miller/GitHub/DAnanodegree/Class3_DataAnalysisWithR/C3L3_ExploreOneVariable/pseudo_facebook.tsv', sep = '\t')
names(pf) # Column Names
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
# install.packages('ggplot2')
library(ggplot2)

names(pf)
qplot(x = dob_day, data=pf) + 
  scale_x_discrete(breaks = 1:31)

# ggplot(aes(x = dob_day), data = pf) + 
#   geom_bar() + 
#   scale_x_continuous(breaks = c(1,31))
```

***

#### What are some things that you notice about this histogram?
Response: Lots of people with DOB on the first of the month, whereas the last day is quite weak.

***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}
# Make a histogram for each month of the year
qplot(x = dob_day, data=pf) + 
  scale_x_discrete(breaks = 1:31) + 
  facet_wrap(~dob_month, ncol = 3)

# Facet_wrap and facet_grid
```

#### Let’s take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x = friend_count, data = pf )

# Or you can use this: 
# ggplot(aes(x = friend_count), data = pf) +
#   geom_histogram()
```

#### How is this plot similar to Moira's first plot?
Response:

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}
qplot(x = friend_count, data = pf , xlim = c(0,1000))

# Using a layer instead of parameter
qplot(x = friend_count, data = pf ) + 
  scale_x_continuous(limits = c(0,1000))
```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50))
```

### Split between male and female
```{r Faceting Friend Count}
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf,!is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender, ncol = 2)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count,pf$gender,summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:

***

### Tenure
Notes:

```{r Tenure}
qplot(x = tenure, data = pf, binwidth = 30,
      color = I('black'), fill = I('#099DD9'))

# The parameter color determines the color outline of objects in a plot. 
# The parameter fill determines the color of the area inside objects in a plot. 
# You might notice how the color black and the hex code color of #099DD9 (a shade of blue) are wrapped inside of I(). The I() functions stand for 'as is' and tells qplot to use them as colors. 

ggplot(aes(x = tenure), data = pf) + 
   geom_histogram(binwidth = 30, color = 'black', fill = '#099DD9')
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x = tenure/365, data = pf, binwidth = .25,
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1,7,1), limits = c(0,7))

ggplot(aes(x = tenure/365), data = pf) + 
   geom_histogram(binwidth = .25, color = 'black', fill = '#F79420')
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x = tenure/365, data = pf, binwidth = .25,
      xlab = 'Number of years using facebook',
      ylab = 'Number of years in sample',
      color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(breaks = seq(1,7,1), limits = c(0,7))
```

***

### User Ages
Notes:

```{r User Ages}
qplot(x = age, data = pf, binwidth = 1,
      xlab = 'Age of Users',
      ylab = 'Count of Users in Bin',
      color = I('black'), fill = I('#F79420')) + 
  scale_x_continuous(breaks = seq(0,113,5))

ggplot(aes(x = age), data = pf) + 
  xlab('Age of Users') +
  ylab('Count of Users in Bin') + 
  geom_histogram(binwidth = 1, fill = '#5760AB') + 
  scale_x_discrete(breaks = seq(0, 113, 5))
```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
qplot(x = friend_count, data = pf)
summary(pf$friend_count)
summary(log10(pf$friend_count + 1)) # Add one to avoid -inf 
summary(sqrt(pf$friend_count))
```

```{r Multiple Plots with Varying Scalse}
# install.packages('gridExtra') 
library(gridExtra) 
p1 <- ggplot(aes(friend_count), data = pf) + 
  geom_histogram(breaks = seq(0,1000,10)) + 
  scale_x_continuous(limits = c(0,750))
p2 <- ggplot(aes(log10(friend_count + 1)), data = pf) + 
  geom_histogram(breaks = seq(0,5,.1)) + 
  scale_x_continuous(limits = c(0,4))
p3 <- ggplot(aes(sqrt(friend_count)), data = pf) + 
  geom_histogram(breaks = seq(0,100,1)) + 
  scale_x_continuous(limits = c(0,50))
grid.arrange(p1,p2,p3,ncol=1)

# scale_x_log10()
# Solution per teachers (BASIC)
p1 <- ggplot(aes(friend_count), data = pf) + 
  geom_histogram() 
p2 <- ggplot(aes(log10(friend_count + 1)), data = pf) + 
  geom_histogram() 
p3 <- ggplot(aes(sqrt(friend_count)), data = pf) + 
  geom_histogram() 
grid.arrange(p1,p2,p3,ncol=1)

# Solution per teachers - Advanced - using ggplot syntax
p1 <- ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram() 
p2 <- p1 + scale_x_log10() # Just transforming
p3 <- p1 + scale_x_sqrt()
grid.arrange(p1,p2,p3,ncol=1)

```

```{r Another Scaling Example}
logScale <- qplot(x = log10(friend_count), data =pf)
countScale <- ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram()+ 
  scale_x_log10()

grid.arrange(logScale,countScale,ncol = 2)
```


***


### Frequency Polygons

```{r Frequency Polygons}
# Original code
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10) +
  scale_x_continuous(lim = c(0,1000), breaks = seq(0,1000, 50)) +
  facet_wrap(~gender)

# No need for facet_wrap as there is color that overlays the gender
# Compare two or more dists at once
qplot(x = friend_count, data = subset(pf, !is.na(gender)),
      binwidth = 10, geom = 'freqpoly', color = gender) + 
  scale_x_continuous(lim = c(0,1000), breaks = seq(0,1000, 50)) 

# Change y axis from raw counts to proprotions
qplot(x = friend_count, y = ..count../sum(..count..),data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, geom = 'freqpoly', color = gender) + 
  scale_x_continuous(lim = c(0,1000), breaks = seq(0,1000, 50)) 

# Where do women overtake men???
qplot(x = friend_count, y = ..count../sum(..count..),data = subset(pf, !is.na(gender)),
      xlab = 'Friend Count',
      ylab = 'Proportion of Users with that friend count',
      binwidth = 10, 
      geom = 'freqpoly', color = gender) + 
  scale_x_continuous(lim = c(350,1000), breaks = seq(350,1000, 50)) 

# GGplot syntax for reference
ggplot(aes(x = friend_count, y = ..count../sum(..count..)), 
       data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')

# Check out where women overtake men?
ggplot(aes(x = friend_count, y = ..count../sum(..count..)), 
       data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) + 
  scale_x_continuous(limits = c(350, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')


```

***

### Likes on the Web
Notes:

```{r Likes on the Web - Which Gender}
# This log transformation really shows where females overtake males
ggplot(aes(x = www_likes), 
       data = subset(pf,!is.na(gender))) + 
  geom_freqpoly(aes(color = gender)) + 
  scale_x_continuous() + # we konw that www_likes is continuous
  scale_x_log10()

# how many www_likes there are in the entire data set for males?
# Which gender has more www_likes in total?

 # tapply splits the data by the second argument, then applies the third argument function to the variable
by(pf$www_likes,pf$gender,sum) # Class Solution 
tapply(pf$www_likes,pf$gender,sum) # Andy soulution

```


***

### Box Plots
Notes:

```{r Box Plots}
# qplot automatically creates a histogram, so we need to pass a geom PARAMETER to tell it to do boxplot
# for box plots - the grouping variable is always X and the continuous variable is always Y.
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot')
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}
# Original Solution - BUT you are removing data points!
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot', ylim =  c(0,1000))

# Another solution - BUT you are removing data points!
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') + 
  scale_y_continuous(limits = c(0,1000))

# Advanced Solution - This affects the actual top of the box - because now the IQR is calculated on ALL the data, instead of a subset of the data!
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') + 
  coord_cartesian(ylim =  c(0,1000))
```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}
qplot(x = gender, y = friend_count, 
      data = subset(pf, !is.na(gender)),
      geom = 'boxplot') + 
  coord_cartesian(ylim =  c(0,250))

# Analyzing the friend count variable, want to split it over gender, and generate summary function
by(pf$friend_count, pf$gender, summary)
# So 75% of females have a friend count below 244 friends, OR 25% of females have friend counts greater than 244
```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}
qplot(x = gender, y = friendships_initiated, 
      data = subset(pf, !is.na(pf$gender)),
      geom = 'boxplot') + 
  coord_cartesian(ylim = c(0,150))

by(pf$friendships_initiated,pf$gender,summary)
```

Response:
I can see by looking at the box plot and the summaries that woman on average initiated more friendships...  The median is just slightly higher (by 5) from 44 for males and 49 for females.  This jives with the mean as well.  Further, we can see that 50% of females have initiated between 19 and 114 friendships.
***

### Getting Logical
Notes: Want to convert a variable that has alot of 0 values to a binary variable (0 False and 1 True)

```{r Getting Logical}
summary(pf$mobile_likes)
summary(pf$mobile_likes > 0) # Get logical output bc used a comparison operator.

pf$mobile_check_in <- NA #This creates a new variable in the data frame with NA values. 
pf$mobile_check_in <- ifelse(pf$mobile_likes > 0, 1, 0)
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)

# My solution
anm = summary(pf$mobile_check_in)
usage = anm[2]/(anm[1] + anm[2])
usage

# Class solution
usage_class_solution = sum(pf$mobile_check_in == 1) / length(pf$mobile_check_in) # Length is the length of the vector
usage_class_solution
```

Response:

***

### Analyzing One Variable
Reflection:
First and foremost, I learned about exploration of data using simple visualizations and general table summaries.  Exploring my data allowed me to further understand the data and generate new thoughts about how to interpret / analyze the data.  
***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!